% ************************** Thesis Abstract *****************************
% Use `abstract' as an option in the document class to print only the titlepage and the abstract.
\begin{abstract}
We live in the age of data where data is being collected 

Information and knowledge is a powerful thing

There is an emphasis on data-driven and evidence based decision making and knowledge from data can lead to better decisions and actions. Having that resource available for the public creates new opportunities for people to be more involved in creating the spaces they live in larger scale of civic participation

Smart cities are a good example of promotion and use of data for improving the lives of citizens. Data helps understand the impact of cities and complex interactions in it. Opening up that data has potential to innovate public services, increase transparency and empower citizens to take action. 

Open data movement and other initiatives promote data publishing and openness around data
there's multiple initiatives from Open Data Chapters to ODI and 

Also the increasing popularity of Citizen Science


The impact of that data is low or non existent
But there's lack of support for communities to use that data.

In all of this there's is a big role of civil society - organisations, groups and actors positioned between the decision-making powers and citizens. Lack of impact on governance and work on engaging the civil society and working on inclusivity. These are due to issues of accessing, interpreting/ contextualising and making sense of open data for usages in local action. Making data relevant to people in the context and place. Lack of support for civil society.
Lack of resources and know-how among civil society.


The value of data in supporting citizen participation in processes of place-making and community building is widely recognised. While the open data movement now permits citizens to acquire governmental data relating to their communities, little to no effort is made to ensure that these datasets are accessible and interpretable by non-professionals. Through a series of community engagements spanning an 18-month period, we co-designed Data:In Place, an open source web tool which supports citizens in accessing, interpreting and making sense of open data. Leveraging visual map-based querying, citizens can access official statistics about their community, interrogate the data, and map their own data sources to create data visualisations. Reflecting on the participatory design process and the designed technology, we provide a framing to make open data work for civic advocacy.

This PhD research explores more inclusive and participatory ways of designing data-driven, people-centred smart city tools that serve the needs of citizens. In contrast to passive participation through data, this research focuses on ways of using data and technology for bringing people together to address issues in the community, and to intervene in the public space. Through different case studies I explore the role of data in civic participation and advocacy, focusing on active participation through data and digital tools that support collaborative sensemaking of that data.

There is a growing public, political, and academic discourse around the idea that data has the potential to empower citizens. In particular, evidence-based policymaking is at the centre of national and regional planning processes. At the same time a shift towards Localism in planning means that while citizens and civic groups are centrally involved in decision-making about their communities, they lack the skills, resources and access to data that might inform their decision-making. This paper explores the need to establish new ways of supporting deliberation and decision-making in local planning. We report a study that explores the role of data in the complex processes involved in consultation events and the broader collaborative processes surrounding them. In doing so we highlight the need for the integration of dialogic forms of participation with other locally produced data, and for this to be shared in ways that position data as a resource for action.


\end{abstract}