%!TEX root = ../thesis.tex
%*******************************************************************************
%****************************** Third Chapter **********************************
%*******************************************************************************
\chapter[SenseMyStreet]{SenseMyStreet - Sensor-Commissioning Toolkit for Communities}

% **************************** Define Graphics Path **************************
\ifpdf
    \graphicspath{{Chapter3/Figs/Raster/}{Chapter3/Figs/PDF/}{Chapter3/Figs/}}
\else
    \graphicspath{{Chapter3/Figs/Vector/}{Chapter3/Figs/}}
\fi

\section{Introduction}
A strong case has been made for evidence based decision-making by using data as resource to make educated and informed decisions. However, there is a disjoint between the people who are affected by these decisions and them also having access to tools and resources to collect data in order to provide the needed evidence for change. To investigate how sensor-commissioning toolkit can support citizen evidencing local concerns and to understand the processes linked to operating a sustainable and transparent sensor-commissioning toolkit, a number of deployments and engagement activities have been carried out in Newcastle Upon Tyne and surrounding areas over the period of three years. This Chapter describes the iterative design process and deployments of SenseMyStreet, a project that has been set up together with Urban Observatory – UK’s largest urban data sensing network. Sensor-commissioning toolkit enables citizens and communities to commission environmental monitors and determine where they will be located on their streets, collecting evidence to inform local issues. SenseMyStreet toolkit was officially launched in the summer of 2017. Since then over 50 people have signed up, 7 groups of people have done environmental sensing with hand-held monitors loaned from the toolkit in their commutes to explore issues in their local area and two communities have commissioned stationary sensors to be deployed in their neighbourhood. Through constant engagement activities with the communities and feedback from online systems, the toolkit went through an iterative design process to involve into an operational citizen sensor-commissioning toolkit.

This chapter sums up the process of setting up the toolkit and the development and deployments carried out by the author and the technicians from the Urban Observatory. The data will also include the fieldwork conducted by the author which studied different groups using the toolkit and appropriating the data gathered from the environmental monitors. The results from this project will be used to analyse the methods used in the design process and the overall challenges and opportunities associated with developing a transparent sensor-commissioning toolkit, which allows people to collect, share, explore and purpose data collected using high precision environmental sensors. Furthermore, the toolkit continues to operate as part of Urban Observatory at Newcastle University to give communities and citizens the power to gather data relevant to issues that are important to them at a very local scale.

\section{Background/Related work}
Participation in smart city

Data-driven

Participatory sensing, VGI

Citizen science (Hakalay) from crowdsourcing to extreme citizen science (group projects)

Pittsburgh air pollution monitoring, aircast

Tools?? Maybe don’t focus on tools because they are just tools.

Community organising

Community networks - (John M Carroll and Mary Beth Rosson. 2003. A Trajectory for Community Networks. The Information Society 19: 381–393.)

\section{The process/Design of the sensor commissioning toolkit/Designing sensor commissioning toolkit}

Everyone needs their space to live, work and get around doing their everyday activities. However with more than half of the world's population now living in cities it is becoming increasingly more difficult. It is also estimated that the urbanisation will continue and almost 90\% of they world's population will live in densely populated urban environments by the year 2050. As the space in the centres become more of a deficit, cities become dispersed and more people inhabit the outskirts of urban areas. Although people live on the edges of downs and cities, they often work or have to travel because of work to the centres. This creates increased volumes of traffic which consequently creates problems of congestion and air quality. According to UK government's department of transport the total number of vehicles in Great Britain has increased every year since the Second World War with an average growth of 640 thousand per year. This means that cities are faced with a multifaceted problem that requires the collaboration of policy-makers, city architects, transportation and city planners, public health officials, but also citizens. 

In July 2017 the Department for Environment, Food and Rural Affairs and Department for Transport announced its plan for tackling roadside nitrogen dioxide concentrations in the UK. This was a direct response to the illegal levels of air quality in 45 of local authorities in the UK. Among them were Gateshead Metropolitan Borough, North Tyneside and Newcastle upon Tyne which are the focus areas of this case study. The local authorities have to come up with a combined task force to come up with proposals and solution to tackle these illegal levels of air pollutants. There are already two 'Air quality Management Zones' is the Newcastle upon Tyne which have been under constant monitoring from 1995 since the release of the Environmental Act. Up to this date, the local council has taken steps to promote alternative means of travel and invested in infrastructure for active travel through the The Cycling and Walking Investment Strategy (CWIS).

One of the key drivers for [city]’s smart city plans is its poor air quality. For many cities in the UK, [city] included, ‘Air Quality Monitoring Zones’ have been set up by local authorities in an attempt to monitor and reduce air pollution. The city contains two Air Quality Monitoring Zones, where air pollution levels exceed maximum levels prescribed by the UK government [reference omitted for review] and the local authority is obliged to explore remedial measures, but also several other areas where air quality is in need of improvement (including the inner city neighbourhood we worked in). So far, the local authority has put in place Air Quality Action Plans for the two zones, implemented transport and placemaking measures in nearby neighbourhoods to promote walking and cycling over car use, and explored the feasibility of more controversial measures such as restricting access to the city centre for more polluting vehicles [reference omitted for review].


Parallel to this, residents in several neighbourhoods have taken the proactive step of setting up action groups. These groups broadly seek to draw attention to the negative impacts of air pollution and car dependency and promote walking and cycling. In the neighbourhood we worked in, a mixed and diverse area two miles from the city centre and popular with students and young professionals, a handful of residents helped set up a group called [name of community group redacted]. Among their objectives are campaigning for cleaner air, securing improved walking and cycling infrastructure, and advocating for open data about air pollution to inform travel choices, along with increasing public awareness around all these issues. In particular, they express concerns about the effects of air pollution on young children, who they argue are particularly vulnerable. One of the authors was already working with the group to help them collect air pollution data using portable, handheld air quality monitors that formed part of [a citizen sensor commissioning toolkit]. This toolkit, developed in collaboration with [the host university of this author], comprises a selection of professional-grade environmental monitoring tools for appropriation by citizens to sense their environment and commission permanent sensors in parts of the city where they identify problems. Whilst the sensing tools were not specifically designed for use in this way by the manufacturers, they were selected by the designers of the toolkit for their portability and ease of use, and come with simplified instructions rewritten for a lay audience. 
In January 2018, [the community group] reached out to us to seek our support for conducting public engagement around air pollution with school children through using this toolkit. Initially, the group was just interested in involving children in gathering air quality data using the handheld monitors. But we wanted to go further than this: we considered this to represent a unique opportunity to begin open-ended conversations with children around the potential for ‘sensing’ the city – the senses being a primary gateway to exploring the environment [28]. We also saw this as a chance to involve local children in a collaborative investigation into their urban environment, where air quality could form one part of a bigger mix. In doing so, we hoped to offer local children the chance to experiment with using the sensing tools within the scope of its technical affordances, but also think about how they may be able to use these technologies to improve conditions in their neighbourhood. 
We agreed to collaborate with the group on a pilot engagement that would comprise three workshops and take place in a local primary (elementary) school, with a view to conducting more research engagements in other local schools in the near future. This primary school is situated very close to the neighbourhood centre and over half of its students live in walking distance. We met with the headteacher of the school, who drew our attention to issues that the school faces with being on a busy road, where air quality, noise pollution and traffic all causes problems. At the recommendation of the headteacher, we worked with a class of eight and nine year olds (fifth graders). Meeting afterwards with their class teacher, we worked together to refine our workshop designs to ensure that they were appropriate for the children’s needs and abilities and that they complement their learning in Geography and Science.


Although issues with air quality is not the sole focus of the study around citizen participation in the smart city it is definitely one of the key drivers for the local authority and the citizens effected and involved in the study.



and will help bridge the gap in the journey towards zero emissions by 2050. 

% Background about the city and the polices
Council/Government work The Cycling and Walking Investment Strategy (CWIS)

% stats about the car usage and city becoming too small to everyone to get around

% Background and justification for taking up the project

Being a part of six national urban observatories in the United Kingdom, Newcastle's Urban Observatory aims to provide a resource for understanding the impact the city and the complex interactions in it through sensing and data collection. The observatory operates as openly and transparently as possible to provide the best public access to environmental information from the city and do it the way in which promotes the information being used by a wide range of stakeholders - environmental scientists, city officials, universities, businesses, but also including regular citizens to promote initiatives that involve community-led investigation. Despite of the best efforts of the project to make data accessible and usable for everyone the data was mainly used by individuals or groups of expert users thus not living up for the slogan of the observatory - "Data for everyone". The questions that surfaced where whether the public actually cares about the environmental data or is there a need for a different approach of citizen engagement?

The aim of Urban Observatory Increasing efforts of engaging citizens with the smart city 

The emphasis is on initiatives where citizens are involved in problem definition, data collection and analysis. We are investigating the possibilities around re-configuring smart cities to be led by the real needs and desires of citizens, rather than dictated by available technologies and the assumptions of decision makers. 

The aim is to include public access to environmental information and the way in which the information is used by a wide range of stakeholders, including citizen science and in particular applications that involve community-led investigation.
But does the public actually care about environmental data? 

development and use of participatory GIS and mapping, and Human-Computer Interaction (HCI) for geospatial technologies

% Workshop in Gosforth (Beyond pins on the map)


% Planning proposals and changes (Streets for People, Cycle ambition fund)

% Activist groups informing (Groups of concerned citizens) 

% Community support

% Press interest, SPAN, available resources


When did the project start and who initiated it?

SenseMyStreet is my PhD (https://digitalcivics.io/) project developed in collaboration between Open Lab (https://openlab.ncl.ac.uk) and the Urban Observatory (http://www.urbanobservatory.ac.uk/) at Newcastle University (https://www.ncl.ac.uk/). We started with the project in summer 2017 with running workshops around usage of data to advocate changes in local communities. We learned that people are interested in environmental data but often in places that it is not available and there’s no plans from the decision-makers to collect it. From their on we established SenseMyStreet that enables people commission scientific grade environmental from the Urban Observatory and locate them in places they are interested in, collecting data to inform or explore issues they are having. The project itself consists of multiple steps where people identify the issues, use hand-held sensors to do their own investigation and then focus on areas that need more data collection and installing stationary sensing equipment. More information is available at https://sensemystreet.uk/about/. 

How are you communicating the project to the citizens and recruiting them?

At first the we communicated information about the project in different community meetings which we attended and also a lot of local councils stakeholder meetings about the city. A lot of information is passed via word of mouth and people pointing others to the website or directly to us. Also I have worked with different community organisations and local interest groups for in the area for over three years now and have established a good relationships with them.


Have you any information on new initiatives there has been made upon the data collection?

When we started there’s was one established community/pressure group that worked on influencing the decisions about the built environment in their local area - http://spaceforgosforth.com/. As we started our work people have now organised in different areas in Newcastle and started their own local groups from the example of the first - https://www.spaceforheaton.com/,  http://www.spaceforjesmond.com/ and others. Aim for these groups is to engage their local area with the issues and activities that people can do change them — one of them drawing attention to the air quality and issues with traffic. Through these groups we have a wider range of engagement. Also they are all collaborating with each other and sharing their knowledge.

Have you any information on CO2 levels or improvement of air quality?

This is a more complicated question to answer. A side note that the main concern has shifted from CO2 to NO2 emissions from diesel cars which are very harmful to people’s health. UK government published their policy around tackling roadside NO2 in 26 July 2017 which requests a response form local governments (including Newcastle council) to respond with their local plan to reduce pollution in the city. UK government has been sued multiple times for not addressing the illegal levels of air pollution in different local authorities - https://www.clientearth.org/government-loses-third-air-pollution-case-judge-rules-air-pollution-plans-unlawful/. In Newcastle there’s 2 air quality management areas (https://www.newcastle.gov.uk/environment-and-waste/pollution/air-pollution/monitoring-air-quality) which are the main concern for the government. ATM the local council is working with the translation providers, city planning officials, University (Including the Urban Observatory) and other experts to look at different solutions to improve the air quality in those areas mainly. However they are also trying to consult with the wider public around the issues of air quality, including the people who are actively involved with the SenseMyStreet project. The local government has to publish their proposed solution in March 2018. However a lot of people are worried because some of the other areas which are not in air quality management areas are still getting high values of pollution and if there’s going to be some change that will cause people to change their movement patterns the other areas can have a bigger impact. As we don’t know the plan at this moment I can’t speculate on what will happen. There needs to be a holistic approach to these things though including input from everyone - citizens, environmental sensors, experts, researchers and decision-makers (government). Also we need to look at things like social inclusion — sometimes making the environment better and air cleaner with clean air zones (taxing people in specific areas) will make people’s, who are from deprived areas, lives more difficult. 
However the only way to tackle the air quality issue is to go directly to the sources — ourselves. So it’s more about raising awareness in local communities and using the data to start conversations and draw people’s attention to the issue. The groups are working with council to deliver new schemes for active travel - https://www.newcastle.gov.uk/parking-roads-and-transport/cycling/streets-people-1, but the data is more useful for engaging the community and working on nudging people to change their behaviour.

Hope this answered some of your questions. I’m in the process of writing a publication about the project as well which will have more details in it. I will definitely let you know when I have something to share.

¥	Identifying the need for data (Discovery Phase). Through conducting interviews and observations of existing groups map out how they are using/not using data. 

¥	Making data and technology accessible (Design/Develop Phase). Undertaking a co-design process to make platforms and tools that respond to the needs of citizens. 
Starting with paper based methods and then moving to digital.

¥	Meaningful presentation of data (Deploy Phase). Ascertain what visualisations and formats of data are more accessible and useful for the sensemaking processes in different contexts. 

¥	Taking collective action by purposing data (Sensemaking Phase). Supporting the processes of sensemaking and assessing the use of data for decision-making and developing new solutions. 



\section{SenseMyStreet Toolkit}

Describe the process step by step