% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

I have been fortunate to be guided by amazing tutors and supervisors over the years of my studies. Thanks to Prof. Dr. Patrick Olivier for giving me the opportunity, Prof. Dr. Thomas Plöetz for guiding me in the early days of the PhD, Dr. Adrian Clear for helping me choose my academic path, Prof. Emer. Dr. Pete Wright for helping me think like an academic and for our stimulating discussions over the years, Prof. Phil James for pushing me to get stuff done and supporting me in my decisions, to Dr. Kyle Montague for being a never ending source of advice and for listening to me go off on tangents in the last phase of the thesis. Special thanks to Dr. Jan Smeddinck for all the support and guidance for my work.

I would love to thank all my colleagues (past and present) in Open Lab who helped and inspired me over the years. Thanks Andy G., Tom N., Rob A., Ed J., Seb M., Dave C., Shauna, Dalya, Samantha, Clara, Vasilis, Nils, Cassim, Chris E., Rob C., John V., Tom B., Clement, Dan J., and many others. Thanks to the amazing team at the Urban Observatory for keeping up with my parachute working. Thanks Jennine, Luke, Neil, Tom, Dave P., Richard and Rich, and Lindsey.

To my fellow CDTs and soon to be doctors -- Reem, Helen R., Nataly, Jen, Janis, Delvin, Jay, Colin, Dan H., Dan R., Stuart N. -- thank you for taking this journey with me, look at us now! Also, thanks to the extended CDT and PhD family -- Ian, Emma, Sean P., Tom M., Lydia, Andy D., Zander, Tim, Euijin, Tagreed, Gerry and others.

Huge thanks to all my friends, whether close to home or across seas and mountains. Thanks Maarten for becoming my friend when I needed it the most and showing me life outside Estonia. Thanks to Kristjan, Raiko and Sven for the support from afar.

Thank you to my parents, especially my mum for putting a roof over our heads and helping me start off my education and adult life. Suur aitäh teile! To my big sister, thank you for your encouragement and for inspiring me to pursue a university degree. I hope I can repay the favour with this one. Also huge thanks to my new family members, Ann and Stuart, who have showed me love and support throughout this journey. Special thanks to my new sister Helen, for all the encouragement, belief and support.

Finally, most importantly I want to thank Antonia, my love, for all of your understanding and patience. Thanks for always being there for me and helping me overcome all the obstacles, self-doubt and fears. I would not have been able to do it without your never ending support and love.

\end{acknowledgements}
