%!TEX root = ../thesis.tex
%*******************************************************************************
%****************************** Eight Chapter **********************************
%*******************************************************************************
\chapter{Conclusions}
\label{chap:conclusions}

% **************************** Define Graphics Path **************************
\ifpdf
    \graphicspath{{Chapter8/Figs/Raster/}{Chapter8/Figs/PDF/}{Chapter8/Figs/}}
\else
    \graphicspath{{Chapter8/Figs/Vector/}{Chapter8/Figs/}}
\fi

\epigraph{I had nothing to offer anybody except my own confusion.}{--- \textup{Jack Kerouac}}


This Chapter summarises presents contributions of the research to multiple fields of Human-Computer Interaction (HCI), Information Systems (IS) engineering and Community Informatics (CI), and concludes by giving short answers to the research questions research questions introduced in the \hyperlink{questions}{Introduction} Chapter.

\section{Contributions}

\subsection{Human-Computer Interaction}
Despite the growing body of research in HCI focusing on the design of inclusive technologies for civic participation, there is lack of research in understanding what enables citizens to become active users of data and technologies for participating in decision making. This thesis establishes an understanding what makes data useful (i.e., effective use) for non-expert citizens and communities, contributing new knowledge of the relationships between data, communities and civic participation, advocacy and action. Through participatory action design research, which is common approach for studies in HCI, and applying engineering sprints to develop and deploy technologies enabled to actively respond to the community problem solving activities. Furthermore, by analysing the way people made use of community data sources uncovered new understandings of civic advocacy and action, which were condensed as Citizen Advocacy Framework (CAF) helping to discuss advocacy efforts and plan for future activities of local action by citizens (Chapter \ref{chap:sensemystreet-eval}). However, the analysis and evaluation of these technologies also revealed the temporality of these solutions responding to the ever-changing community and their activities. Building on the community commissioning research \citep{garbett_designing_2017}, this thesis has extended the digital civics approach \citep{olivier_digital_2015,vlachokyriakos_digital_2016,asad_tap_2017} with the inclusion of designing for building community capacity as a step towards the successful transfer of ownership and the adoption of community information resources (Chapter \ref{chap:sensemystreet-eval} and \ref{chap:cdi}). Additionally, that includes presenting a deeper understanding of the role the expert professionals and instinctual designer or researchers play in this.

\subsection{Information Systems}
Taking a human-centred design approach to designing and developing community information systems this research has documented the design processes and also provided two open-source platforms for IS. Firstly, the SenseMyStreet toolkit (Chapter \ref{chap:sensemystreet}), which is a set of tools and platforms that enable communities use scientific grade environmental monitors to produce and commission data about their neighbourhood. Secondly, the Data:In Place (Chapter \ref{chap:datainplace}), open-source web platform built to support citizens in accessing, interpreting and using open data for purposes of civic advocacy and action. Furthermore, through the community problem solving activities and the user-centred design processes, the thesis has contributed new ways how to accommodate more social agendas into the design and development of future ISs.

\subsection{Community Informatics}
This thesis has provided a new model of Community-Data Interaction (CDI) for the field of CI, that helps discuss and design community resources for knowledge making and civic action through the effective use of data by citizens (Chapter \ref{chap:cdi}). CDI extends the view on design of CI tools from focusing only on the access and instrumentation of control \citep{stillman_community_2009}, to looking at how technologies could be used also to communicate ideas, share knowledge, create community links and increase capacities for effective use of data by communities. Thesis has also contributed findings from a practical use of CDI as technical solution and social facilitation methods that support communities making effective use of data for civic advocacy and action (Chapter \ref{chap:cdi_in_action}). This provides practical examples that the key elements of CDI can take, enabling researchers better adopt the model for the further research in CI and CDI.

\section{Answering the Research Questions}
\subsubsection{What is involved in citizens creating actionable knowledge using data?}
The assumption of lack of skills, access or tools to use data being the only contributes to communities not leveraging data is not completely true. However, what was learnt from the research is even when building better systems for access and use of data it does not necessarily increase data's actionability by communities. Nor providing new ways for citizens to actively produce or commission data will not guarantee the effective use of that data by citizens. The usefulness of data for citizens is dictated by the activities that they plan to do and things they want to achieve. Citizens need a purpose for what they use the data for, only then then it becomes useful to them and can be put in action. Furthermore, those purposes need to be integrated early into the the investigation and issue enquiry process. 

\subsubsection{How do we build and configure tools and processes to support knowledge making and enable effective use of data by citizens in civic advocacy and action?}
In order to enable communities to effectively use data for local benefit we need to move away from building tools just for access and use of data, and start looking at ways we can build tools for social facilitation around the use of data. That includes providing training, methods and digital tools for helping people articulate their concerns, communicate and discuss ideas, understand what resources are needed and plan steps for action. Furthermore, some of these tools are already used by communities for some other purposes, which can be reconfigured to serve the community problem solving agendas. Indeed, there is a need for increased capacities for communities and either transfer of or the support of data science skills. However, that does not mean coming in and doing the work for the communities. In order to sustain the engagement and use of community information resources and solutions, the community needs to be able to take ownership and through increasing community links be able to adopt to constantly changing social economy.